---
title: Data Quality Report
nep: `r ccd@nepisodes`
np: `r max(ccd@infotb$pid, na.rm=T)`
site: `r s <- unique(ccd@infotb$site_id); s[s!="NA"]`
admt: `r min(ccd@infotb$t_admission, na.rm=T)`
disct: `r max(ccd@infotb$t_discharge, na.rm=T)`
dbupdate: `r max(ccd@infotb$parse_time)`
fulldb: `r dbfull`

---

    
```{r, echo=FALSE, warning=FALSE}
item2d <- c("h_rate", "bp_m_a", "bp_sys_a", "lactate_abg", "bilirubin",
            "fluid_balance_h", "pao2_abg")

new.config <- function(items) {
    conf <- list()
    for (i in items) conf[[stname2code(i)]] <- list()
    conf
}
if (!exists("cctb"))
    cctb <<- create.cctable(ccd, conf=new.config(item2d), freq=1)$torigin

suppressWarnings(library(pander))
panderOptions("digits", 2) 
panderOptions("table.split.table", Inf)
demg <<- sql.demographic.table(ccd)
```



# Data Summary
```{r, echo=FALSE}
dp <- total.data.point(ccd)
np <- max(ccd@infotb$pid, na.rm=T)
```

This database contains **`r ccd@nepisodes`** episode data from 
**`r length(s[s!="NA"])`** sites. Based on the NHS numbers and
PAS numbers, we can identify **`r np`** unique patients, among which the earliest
admission is `r min(ccd@infotb$t_admission, na.rm=T)` and the latest discharge time is
`r max(ccd@infotb$t_discharge, na.rm=T)`. There are `r dp `
total data points found in the current database, which makes in average
`r round(dp/np)` per unique patient. 


## Site reference
```{r, echo=FALSE}
pander(as.data.frame(site.info()[, 1:3], style="rmarkdown"))
```

## The original XML files and parse information
```{r, echo=FALSE, results="asis"}
fs <- file.summary(ccd)
pander(as.data.frame(fs[, c("File", "Number of Episode", "Sites"), with=F]))
```

```{r fig.width=15, fig.height=10, echo=FALSE}
xml.file.duration.plot(ccd)
```
```{r fig.width=15, fig.height=10, echo=FALSE}
xml.site.duration.plot(ccd)
```
\newpage

```{r, echo=FALSE, results="asis"}
table1(demg, c("SEX", "ETHNIC", "LOCA", "DIS", "HDIS"))
```
# Completeness
## Demographic data completeness
```{r, echo=FALSE, warning=FALSE, results="asis"}
demographic.data.completeness(demg)
```
## Sample period of time-wise data
This is the section to show the completeness of some key time-wise data, e.g.
physiological data. The completeness of data is measured by sample period. 
The sample period $P$ is defined as the ratio of the total
admission hour $T$ to the number of valid hours $D$ where data can be found.
$$P = \frac{T}{D}$$




```{r, echo=FALSE, warning=FALSE, results="asis"}
samplerate2d(cctb)
```



\newpage
# Data Distribution
## Demographic
```{r, echo=FALSE, results="asis", warning=FALSE}
demg.distribution(demg, c("HCM", "apache_score"))
```


## Physiological & Drugs
```{r, echo=FALSE, results="asis", warning=FALSE}
physio.distribution(cctb, item2d)
```

