---
title: "Introduction to CCHIC critical care data"
author: David Perez Suarez & Sinan Shi
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introduction to CCHIC critical care date}
  \usepackage[utf8]{inputenc}
---

CCHIC data strongly impacts the design of cleanEHR. Complex, heterogeneous, and
high resolution longitudinal data is a trend of EHR analysis, which takes
advantage of the ever more sophisticated statistical techniques and growing
computation capability. CCHIC is a representative example database of such
kind. It records 263 fields including 154 time-varying fields of patients
during their stay in intensive care units across five NHS trusts in England.
The recorded variables include patient demographics, time of admission and
discharge, survival status, diagnosis, physiology, laboratory, nursing and
drug. The latest database contains 22,628 admissions from 2014 to 2016, with
about 119 million data points (~6k per patient).

The anonymised data subset can be obtained [here](https://form.jotformeu.com/drstevok/cchic-end-user-license---cleanEHR). 
The selected number of researchers can get the access to the identifiable data 
[UCL IDHS](http://www.ucl.ac.uk/isd/itforslms/services/handling-sens-data/tech-soln).

### Episode and ICU stay timeline
We introduced the concept of 'episode' as the fundamental entity of EHRs, which
comprises all the data being recorded during the ICU stay. Each episode also
contains the demographic information of the patient, ward transferring origin
and destination within a hospital and diagnosis information. It allows us to 
link episode data from a single patient across the entire multi-centre database. 
The key dates and times are recorded as follow. 
<img src=graphs/cchic_timeline.png width=480 height=360 />
```{r, message=FALSE, warning=FALSE}
library(cleanEHR)
data("sample_ccd")

# Extract all non-longitudinal data (demographic, time, survival status, diagnosis)
dt <- ccd_demographic_table(ccd, dtype=TRUE)
```
`ccd_demographic_table` function returns a table of all non-time-varying fields alongside with 
several derived fields -- the fields that are not directly recorded in the original data. 
Each row of the table is a unique admission, and every column is a non-time-varying data field. 

* `pid`: unique patient ID derived from NHS number or PAS number.
* `AGE`: date of birth - unit admission time
* `spell`: see *Spell*

```{r}
print(dt[1:3, ])
```

Data missing are caused by many reasons. We have to understand that in such a large database, 
the data quality varies. In the anonymised dataset, data can be missing due to security reason. 
Missing data are either NULL or "NA" depending on the field data type. 

### Demographic data
Every patient in England has a unique NHS number and PAS (Patient Admission
System) number. These can be used to identify a unique patient. Other
demographic information can also be found in the CCHIC dataset such as age, sex,
GP code, postcode and so on. Most of the demographic data will be removed,
pseudonimised, or modified in the anonymised dataset to protect the patient
privacy. 


### Spell
Ward transferring within or beyond ICUs are counted as different episodes respectively. 
In some research, one may be interested in looking at the sickness development and the treatment 
history beyond each ICU stays. A spell includes number of episodes from a
unique patient which occured in a user defined period One can link episodes by
spell ID. 
```{r}
head(ccd_unique_spell(ccd, duration=1)[, c("episode_id", "spell")])
```

### Diagnosis data
Instead of using free text, we adopted [ICNARC diagnosis](https://www.icnarc.org/Our-Audit/Audits/Cmp/Resources/Icm-Icnarc-Coding-Method) 
code system to record the diagnosis. The full ICNARC code is a five digit number separated 
with dots. From left to right each digit represents a higher category of diagnosis. Due 
to the privacy concerns, in the anonysmised dataset, last two digits will be removed. 
One may use the function `icnarc2diagnosis` to look up the diagnosis code.
```{r}
icnarc2diagnosis("1.1")
icnarc2diagnosis("1.1.4")
icnarc2diagnosis("1.1.4.27.1")
```

### Longitudinal data
CCHIC measures 154 longitudinal data. The full list of longitudinal data are shown below. 
<img src=graphs/item_ref.png width=480 height=360 />

We can also easily plot the data from a single selected admission. 
```{r, fig.width=10, fig.height=11, out.width='700px', results='hide', message=FALSE, warning=FALSE}
plot_episode(ccd@episodes[[7]], c("h_rate",  "bilirubin", "fluid_balance_d"))
```
